Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-11-26 paper/paperV3.tex
Second revision

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR CNRS 6004, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR CNRS 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
\maketitle

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes (pleasantness) to a sufficient degree of accuracy through a comparative study with state of the art approaches.
\end{enumerate}

Section~\ref{sec:data} presents the generation procedure for a large corpus of simulated sound scenes, its validation through a listening test in the context of a study on perceived soundscape quality, and a study of available acoustical indicators for the automatic annotation of source activity. In Section~\ref{sec:pred} a deep learning architecture that performs multi-label sound source recognition at relevant time scales is trained and evaluated. Section~\ref{sec:app} then presents an application of this model to the prediction of pleasantness in urban environments.

\section{Corpus}
\label{sec:data}

\subsection{Generation procedure}
\label{sec:data_corp}

Considering machine learning techniques to detect the presence of sources requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed. This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}. The corresponding isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. \textit{simScene} generates original scenarios using the following information:

@@

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

This information is reused to extract statistics on background and event sources activity, which are then used as inputs by \textit{simScene}. Though, adjustments are made on the variance of the considered properties to extend the range of covered scenarios. Furthermore, since voice events in the isolated samples database consist mostly of read English recordings, the voice event mean event-to-background ratios are reduced for all ambiances to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes. At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.

@@

The generation procedure is the same for both datasets. Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets. To do so, the isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, the two corpora differ by the scenarios and isolated samples used in their generation.

\subsection{Perceptual experiment}
\label{sec:data_exp}

A listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model. Given the aim of this test, a separate corpus is constructed. This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.

First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a third isolated samples database to ensure independance from the deep learning datasets. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. Figure~\ref{fig:tvb_pres} shows the distribution of the 75 selected extracts. Playback sound levels are realistically drawn for each scene as described in Section~\ref{sec:data_corp}, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.

\begin{figure*}[th]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with scenarios generated by \textit{simScene}.}\label{fig:tvb_pres}
\end{figure*}

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

During the test, participants are asked to evaluate sound scenes using 8 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then loudest of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to listen to the full extract and to answer all questions before being allowed to proceed to the next scene.

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. The calibration of the headphone was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the sound pressure at the entrance of a blocked ear canal. To do so, small DPA 4060 microphones have been set at the entrance of the ear canals of a human subject~\cite{moller1992}. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 subjects, the incomplete block design is not perfectly balanced). All participants gave written consent prior to the experiment, and evaluations were further anonymized.

\subsection{Simulated corpus validation}
\label{sec:data_val}

The effect of the \textit{simScene} generation procedure on perception is first investigated. To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scenarios. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

\begin{table*}[t]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects for active individuals (simulated scenes) are projected onto the principal components space for and represented in Figure~\ref{fig:pspace_sim} as dots. The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space and represented as crosses. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the diversity of scenarios created by the scene generation procedure.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
%\usepackage{tablefootnote}
\usepackage{threeparttable}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR CNRS 6004, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR CNRS 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
\maketitle

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{All statistical analyses are done with Matlab. Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes (pleasantness) to a sufficient degree of accuracy through a comparative study with state of the art approaches.
\end{enumerate}

Section~\ref{sec:exp} presents the generation procedure for a corpus of simulated sound scenes and its validation through a listening test in the context of a study on perceived soundscape quality. In Section~\ref{sec:pred} a larger dataset is constructed and automatically annotated, and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Section~\ref{sec:app} then presents an application of this model to the prediction of pleasantness in urban environments.

\section{Listening experiment}
\label{sec:exp}

\subsection{Stimuli}
\label{sec:exp_stim}

\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/parcours.png}
    \caption{Noise map of the path and 19 locations of the soundwalks presented in~\cite{aumond2017}. Sound levels (Leq in dB SPL) were measured at each of the 19 locations and interpolated to obtain a sound level estimations along the path.}\label{fig:path}
\end{figure*}

A listening test is conducted in order to validate the perceptual correspondence of simulated scenes with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model. To do so, a corpus is constructed that contains 100 sound scenes, including 75 simulated scenes, 6 recorded scenes for reference, and 19 replicated scenes that correspond to simulated reproductions of recorded scenarios.

The 75 simulated and 19 replicated scenes are generated using the \textit{simScene} software, an open source library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. \textit{simScene} can be used in two main modes. First, the \textit{generate} mode is used to create new scenarios. It is controlled by the following information for each source in the considered taxonomy:

@@

\textit{simScene} samples these distributions to generate original scenarios with background and event annotations, then simulates sound scenes as the composition of examples in the isolated samples database, with source type, event-to-background ratios and event onsets given by the generated scenarios.

The second mode is the \textit{replicate} mode, which directly takes as input a scenario with background and event source type, event-to-background ratio and onset-offset information. In this mode, \textit{simScene} only samples the available isolated samples database to simulate a sound scene that replicates the given scenario with different acoustic content.

In this study we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}. The isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The isolated samples database contains 8mn, 17mn and 37mn of background traffic, voice and bird extracts respectively, as well as 30mn, 37mn, and 5mn of event traffic, voice and bird extracts respectively. No background noise is added to these three sources, though some uncontrolled noise is present in the uncleaned extracts composing the isolated samples database. A simulated sound scene is thus a mix of up to six separate channels, each one corresponding to either the background or event activity of one of the sources in the chosen taxonomy.

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~mn in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

This information is reused to estimate the mean and variance of the distributions on background and event sources activity used by \textit{simScene} to generate new scenarios. These distributions are conditional to the four considered ambiances and three source types. The variance of all event inter-onset distributions is manually increased to extend the range of covered scenarios. Furthermore, since voice events in the isolated samples database consist mostly of read English recordings, the voice event mean event-to-background ratios are reduced for all ambiances to improve realism. A fifth \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity. At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.

The 75 simulated scenes of the listening test corpus are generated using \textit{simScene} in the \textit{generate} mode with the extracted distributions as input. To improve the diversity of scenarios considered in the listening test, 200 sound scenes are first simulated, equally distributed among the five ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). A common duration of 45 seconds is chosen due to the limited duration of listening tests because of the fatigue. Though many studies dealing with the environmental acoustic quality are associated with acoustic measurements ranging from a few seconds~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests. To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence for each type of sources (ranging from no presence to 100\% of presence) without redundancy. Playback sound levels are realistically drawn for each scene, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. The path undertaken during these soundwalks is shown in Figure~\ref{fig:path}, with sound levels in dB SPL interpolated from measurements at the 19 locations. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode. The 19 resulting sound scenes are thus simulated, but their scenarios follow those of the reference recordings to the extent of annotation precision. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are scaled using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.

\subsection{Equipment}
\label{sec:exp_equi}

During the test the scenes are played at a given sound level as discussed in Section~\ref{sec:exp_stim}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. The calibration of the headphones was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the corresponding binaural sound pressure. To do so, the following procedure was conducted:
\begin{enumerate}
\item A pink noise generator is set at an arbitrary level and its output voltage is measured.
\item Small DPA 4060 microphones are set at the entrance of the ear canals of a human participant~\cite{moller1992}. The generator is used as input of the headphones, placed over the head of the participant. The voltage at the output of the binaural microphones is measured.
\item The headphones are removed, and the generator is used as input of Genelec 1031A loudspeaker placed at a distance of 1 meter from the participant's head. The loudspeaker's amplification is tuned until the same voltage is outputted by the binaural microphones.
\item The head is replaced by a sound level meter to measure the sound level in dB of the loudspeaker. This corresponds to the binaural sound level produced by the headphones for the considered output voltage of the pink noise generator.
\end{enumerate}
By repeating this procedure for different settings of the generator level, the relation between the logarithm of the generator output voltage and headphones playback sound level in dB is approximated as a linear function. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener, when using the calibrated headphones and a specific sound card configuration for which the relation between the numeric level and output voltage is known.

\subsection{Participants}
\label{sec:exp_part}

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 participants, the incomplete block design is not perfectly balanced). All participants gave written consent prior to the experiment, and evaluations were further anonymized.

\subsection{Procedure}
\label{sec:exp_proc}

\begin{table*}[t!]
\centering
\caption{Mean differences of perceptual assessments (resp. pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point scales (0-10). These attributes are presented in French and translated in this report using standard terminology. The first 5 semantic differential scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then loudest of the recorded scenes (resp. P3 and P15). A random listening order is generated for each participant to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to listen to the full extract and to answer all questions before being allowed to proceed to the next scene.

\subsection{Results}
\label{sec:exp_res}

The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab. The effect of the \textit{simScene} generation procedure on perception is first investigated. To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scene compositions. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point discrete scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for recorded and replicated scenes (n=25) and simulated scenes (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the perceptual spaces based on simulated sounds or on recordings and replicated scenes, do not reveal major differences. Additionally, the assessments averaged on all participants for active individuals (simulated scenes) are projected as dots onto the principal components space in Figure~\ref{fig:pspace_sim}. The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space as crosses. These projections show that the space covered by simulated scenes based on new compositions covers that of the studied real-life environments (recorded and replicated sounds). This further demonstrates the diversity of scenarios created by the scene generation procedure. Additional information on the relation between pairs of recorded and replicated scenes is available on the companion website\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/web/index.html}}.

%Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented using ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

\begin{figure}[th]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_recrep.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_sim2.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes (active individuals) are projected as dots, and recorded and replicated scenes (supplementary individuals) are projected as crosses.}\label{fig:pspace_sim}
\end{figure}

\subsection{Corpus}
\label{sec:pred_corp}

Considering machine learning techniques to detect the presence of sources requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed. This corpus is composed of sound scenes simulated using \textit{simScene} in the \textit{generate} mode with the same procedure as for the 75 simulated scenes in the listening test corpus described in Section~\ref{sec:exp_stim}.

@@

The generation procedure is the same for both subsets, and reuses the distributions extracted from recordings of the GRAFIC project in Section~\ref{sec:exp_stim}. Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets. To do so, an isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is constructed that is independant from the one used in the generation of simulated scenes in the listening test corpus. This database contains 12mn, 49mn, and 73mn of background traffic, voice and bird extracts respectively, and 32mn, 12mn, and 5mn of event traffic, voice and bird extracts respectively. The isolated samples database is split in the same proportions as the two subsets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, the two subsets and the listening test simulated scenes are simulated using three separate, independant isolated samples databases.

\subsection{Acoustical indicators for soundscape description}
\label{sec:pred_inds}

To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:pred_corp}, the activity of sound sources present in each scene must be known. Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source. Several indicators are identified in the literature that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}. These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.
